### 2024-06-16 Paper/main.tex
Updates from Overleaf

    \begin{center}
        \LARGE
        \textbf{A Perfect Storm and the Importance of Favourable First-Nature Geography} \\

@@

    \normalsize
    Is geography destiny? What is the role of first-nature geography in determining prosperity? This paper estimates the effect of randomly removing and introducing favourable first-nature geography to a specific region using a difference in difference design. In 1825 a storm introduced a new navigable waterway, bringing trade and prosperity to the otherwise relatively isolated northwestern Denmark. And 700 years prior, the same event happened in reverse, when a previous channel closed up between 1086 and 1208. The elasticity of geography-induced market access is estimated to be 1.6, corresponding to 24 percent population growth within a generation of the event. Demonstrated mechanisms include trade, fertility, fishing and the rise of manufacturing. The central finding is replicated in reverse in a register of dated archaeological sites. The 1086-1208 closing caused fewer buildings and sites containing coins. The general insight is the same: First-nature geography determines levels of prosperity.

@@

    \footnotesize
    $^*$This project was made possible through generous funding by Independent Research Fund Denmark (DFF – 6109-00123). I want to thank Paul Sharp, Christian Møller-Dahl, Casper Worm Hansen, Max Schulze, Gregory Clark, James Fenske, Mathias Barding, Torben Johansen, Nadja van't Hoff, Kerstin Enflo, Neil Cummins and former PhD student-colleagues (too numerous to mention individually) both at SDU and at LSE for valuable feedback. This project also benefitted from the conscientious research assistance of Andreas Slot Ravnholt. This project was presented at the 2021 and 2022 DGPE workshops, the 2022 EHES conference, the SSEH meeting in 2022, in the LSE Economic History Graduate Seminar series in 2022, the EHS conference in 2023, the End-of-Semester workshop of the Department of Economics 2023 (University of Copenhagen), the Annual CITP conference (Nottingham), the Danish Society of Transport Economics (Copenhagen) and multiple times at the PhD seminar series at the Department of Economics (University of Southern Denmark). I am grateful for the insightful comments and suggestions by participants at these events. This paper also benefitted from improvements suggested by ChatGPT. Important early inspiration for this paper came from the podcast 'Kongerækken' and Poulsen (2019). Finally, I want to thank my fiancee, Maria Fay Courtney Bohr. On our vacation during the COVID-summer of 2020, we stumbled over some fascinating local history resulting in the present paper.  All mistakes are my own. The paper was previously called "A perfect storm and the natural endowments of trade-enabling infrastructure". All code to reproduce all the steps from raw data to the final regressions and the paper itself is available at \\ \url{https://github.com/christianvedels/A_perfect_storm}

@@

\section{Introduction}
In the broadest sense, there are two types of long-run explanations for why some places prosper while others struggle: Institutions \citep{rodrik2004institutions} and geography \citep{Henderson2018satelite}. A predominant consensus suggests that institutions play a primary role, with geography exerting its influence through its impact on the formation of institutions \citep{Acemoglu2001, rodrik2004institutions, Easterly2003, Ketterer2018}. But even among geographical explanations, the evidence to suggest the importance of path-dependence and second-nature geography \citep{Krugman1991, Bleakley2012, Ager2020a} and relegates the role of first-nature geographical fundamentals as being merely important as seeds of future growth and development \citep{Davis2002, Bosker2017, Allen2023}. This paper demonstrates the importance of first-nature geography in its own right. This is done by using a natural experiment which suddenly changed first-nature geography. This in turn caused an increase in trade, population density and both manufacturing and fishing: A large permanent causal effect of first-nature geography on the level of economic activity.

In 1825 a violent storm breached the Agger Isthmus in Northwestern Denmark. An erosion process was initiated and in 1834 a navigable channel had formed. The isthmus had previously separated the Limfjord from the North Sea. But now the region was endowed with convenient access to the open ocean and world markets (see Figure \ref{fig:main_map}). Market towns of the West Limfjord had previously been small and isolated. Now they grew to become relatively important ports of international trade. Similarly, a channel in approximately the same location closed between 1085 and 1208. After the last ice age, Northwestern Denmark was left with a convenient westward channel, used by traders and Viking fleets alike. But it would close due to gradual land-rises - an ongoing process since the last ice age \citep{Christensen2004}. What followed was a long-run stagnation of the West Limfjord region until the geographic circumstances was reversed in 1825.

This paper shows evidence for the effects of these shocks in a difference in difference framework. This is done using historical census data, a large database of Northern European historical trade (the Sound Toll Register) and the densities of archaeological findings. It is shown that the 1825 breach led to an increase in population of 24 percent in 1901. To document the mechanisms, I have collected new data on all Danish occupations (13.5 million observations) in the period 1787-1901. The Danish censuses contain a description of each person's occupation. A new machine learning procedure, \textit{OccCANINE}, is applied to these descriptions to convert them into standardized HISCO codes with high accuracy \citep{dahl2024breaking}. The HISCO-coded census data is then used to estimate the effect on occupational structure. The key finding is that the channel caused both fishing and manufacturing to grow. An important question is whether the improved geography caused growth or simply reallocation of prosperity. Using age, gender and birthplace information it is shown that the population growth came primarily from fertility - not migration. Denmark of the 19th century was in a post-Malthusian regime, where improved living standards cause fertility \citep{Jensen2022, Galor2011}. As such, the evidence of this paper suggests that intrinsic growth is the most likely explanation of the observed effect. That is, improved first-nature geography seemingly causes intrinsic growth - not just reallocation.

For the 11th-century closing, the analysis relies on archaeological data and the closing of the previous channel as an exogenous shock. The time variation comes from the dating of archaeological sites. The geolocation of sites containing coins and buildings is used to construct panels of the probable location of economic activity. A branch within archaeology, Settlement Scaling Theory \citep{Ortman2020}, argues for such approaches to infer historical economic activity, and archaeological evidence has also seen application within economics \citep{Davis2002, Barjamovic2019, Bakker2021Phonecians, Allen2023}. In this paper, I develop a novel method to determine the probability that a coin or building finding was left for later archaeological discovery at any given time. Results show that a one percent change in market access led to a 13 percent decrease in the probability of coin findings being generated. Denmark of the 11th century is very different from its 19th century counterpart. But despite differences in technology, culture, institutions, religion, and most other aspects thought important for economic development, market access played a similar role in the 12th century as it did in the 19th century, providing evidence of (temporal) external validity of the effect of first-nature geography on prosperity.

A separate but related contribution of this paper is in understanding the benefits of infrastructure. Some of the most expensive and extensive projects that a society can undertake are infrastructure projects. Infrastructure changes distances between people. Infrastructure changes the first-nature geography. However, infrastructure investments are often endogenous to outcomes of interest.  If a dramatic increase in e.g. population density is observed following infrastructure, it is difficult to identify whether this is because it was just constructed in an already thriving place. There are different standardized identification strategies commonly applied in the literature \citep{Redding2015}. This paper can be seen as an entirely new way of studying infrastructure change by relying on changes in the natural endowments of it.

@@

This paper proceeds as follows. Section 2 contains an overview of related literature. Section 3 presents the historical background. Section 4 presents the overall empirical strategy. Section 5 presents a general overview of the data used. The remaining sections demonstrate empirical results. Section 6 demonstrates how the shock caused an increase in trade. Section 7 shows how the channel caused the population to grow. Section 8 covers mechanisms. Section 9 demonstrates the effect of the closed-up channel in the 1100s. Section 10 concludes.

@@

This paper enters a debate about the fundamental determinants of prosperity. Is geography or are institutions the fundamental cause? \cite{Acemoglu2001} argue that extractive institutions set up by colonizing Europeans are partially to blame for the current low-income status of many African nations. \cite{rodrik2004institutions} argue that geography plays a limited role if institutions are kept constant. \cite{Henderson2018satelite} use satellite image data to show that geography through trade and agriculture can explain 47 percent of the worldwide variation in economic activity. \cite{Ketterer2018} find that first nature does play a role but institutions are far more important even when a comparison is made within European regions. It would be natural to get the impression, that geography is simply unimportant in its own right. However, what is demonstrated here, is that the Western Limfjord region lacked favourable geography and suddenly got it, and as a consequence, they prospered within a generation.

At this point, it is useful to provide some clarity regarding some terms already used and provide a simple taxonomy of geography and economics. \cite{cronon1992nature} introduced the distinction between first- and second-nature geography, where first-nature geography is everything determined independently of humans, and second-nature geography is all of the geography caused by humans and their interactions. This has since become the standard framework for understanding the interplay of economics and geography in broad terms \citep{Caruana-Galizia_Okubo_Wolf_2021}. Among first-nature geography, we can distinguish between:
\begin{enumerate}
    \item \textit{Things in/on the soil/land}: Soiltype, natural resources, rainfall, etc.
    \item \textit{Geomorphology}:\footnote{Also often called 'physical geography'. But this term is confusing. A bridge is very physical. But it is not what is meant by 'physical geography'. A land-bridge on the other hand is 'physical geography'. Using the term Geomorphology circumvents this confusion.} Waterways, lakes, mountains, etc.
\end{enumerate}

A huge literature is dedicated to second-nature geography in terms of infrastructure (covered below). Among first-nature explanations of the 'what's in the soil'-type, \cite{Nunn2011a, HeavyPlough2016, Dalgaard2020} demonstrate that variation in the productivity for food (the potato, the heavy plough and fish stock respectively) of a particular place caused increased population density. But 'what's in/on the soil'-first-nature also affects other outcomes.  \cite{Decet2023WaterWars} show how a lack of rainfall is associated with an increased level of conflict in modern-day Africa. \cite{WinnersAndLosers2022} demonstrates how land productivity caused inequality, and \cite{clark2014} demonstrate the influence on literacy. Of course, the ground might be endowed with other determinants of growth like coal \citep{ORourkeCoal2021} or peat \citep{CoalPeatPaper}.

On the other hand first-nature \textit{geomorphology} explanations are more sparsely covered with \cite{Diamond1997, Allen2023, Matranga2024} - all related to the neolithic revolution and the emergence of the first civilisations - as the most notable exceptions. Furthermore \cite{Henderson2018satelite} and \cite{Bosker2017} demonstrate that \textit{geomorphology} is correlated with prosperity. Of course, this is often based on the comparison between places with and without certain geomorphological features, which makes the causality contingent on the degree to which you can trust their method for controlling other differences between these places. Achieving a clean identification remains a formidable challenge. Consequently, a survey by \cite{Bosker2022} calls for the exploration of natural experiments to gain insights into causal effects of first-nature \textit{geomorphology} on population growth. The present paper tries to fill that gap.

All of this is related to a central result by \cite{Davis2002}: Even very large disruptions of city populations (the bombs over Hiroshima and Nagasaki at the end of WWII) do not change the long-run outcomes. \cite{Brakman2004Bombs} replicates a similar result for the bombing campaigns in Germany. This result has two types of potentially overlapping explanations: It is either because \textit{first-nature geography} is the fundamental determinant of city population for a given state of the economy, or it is because of path dependence: It is \textit{fundamentals} or \textit{path dependence}.\footnote{\cite{CraftsWolf2014} shows a case for this: Cotton mills grew where the first-nature geography, in the form of available water power, was prevalent. When this was replaced with steam power, the location was persistent due to an interaction of first and second nature geography - some first nature geography (water access) is useful for both water power and steam powered mills, and the investments in buildings and surrounding infrastructure (second nature), might as well be made to good use rather than reallocating it. } In both cases, city sizes will tend to gravitate towards their long-run trend despite of disruptions. But if is due to first-nature geography, then a shock to fthis should have a permanent effect. This what I demonstrate.

That is not to say, that path dependence does not play a role. \cite{Bleakley2012} convincingly documents a case of path dependence. Places which used to be in favourable geography kept being urban centres. An implication of \cite{Krugman1991} is that small initial first-nature geographical differences can accumulate into very different outcomes via second-nature geographical interactions \citep{Caruana-Galizia_Okubo_Wolf_2021}, which is consistent with the observation of \cite{Bleakley2012}. The present paper documents fairly large and permanent effects within a generation when first-nature changes, which firmly demonstrates an important role of the \textit{fundamentals} explanation of the location of economic activity. When favourable geomorphology in the Limfjord region deteriorated, so did the local economy, and when it reverted 700 years later, the economy (and population) increased dramatically. That is, it is not just path dependence.\footnote{Two more related publications are \cite{Ager2020a} and \cite{Cermeno2019}. \cite{Ager2020a} show how a natural disaster can change relative city sizes. And on the other hand \cite{Cermeno2019} show that institutions can create towns in otherwise unfavourable geography.}

The central natural experiment used in this paper is related to \citep{Seror2020Random}. He uses the randomly altered courses of the Yellow River in China to measure the effects of waterways on population growth. The paper finds that market access led to increased population, increased taxes, and in turn reallocation of elites - presumably to avoid the taxes. \citep{Allen2023} uses river shifts and archaeological data in what is today Iraq to demonstrate that river shifts are associated with state formation. In comparison to these two papers, the relatively rich Danish data and the unique natural experiment, allow us to exploit both the 11th- and 19th-century effects of the same event, and especially the 19th-century data, combined with detailed historical accounts, makes it possible to explore the dynamics of adaption to the new opportunities given by favourable geography.

Geomorphology affects market access. That is the fundamental mechanism across most papers mentioned above. As such, we can also compare the present paper to the literature on market access and infrastructure. Infrastructure is second-nature geography, but a special case: It is a deliberate change of the first-nature \textit{geomorphology}. Railways, roads, bridges, canals - even the internet - all change the effective distance between people, which used to be different because of the geomorphology. In general, a large number of papers show how infrastructure shapes everything from land value, health, industrialization, manufacturing and the persistent location of modern-day economic activity \citep{Donaldson2016, Berger2017, Berger2019a, Hornbeck2019, Bogart2019, Zimran2020a, Bogart2022, Hornbeck2024, Hornung2015}. Of course, a central worry is the endogeneity of infrastructure. Infrastructure is often built because a place is already prospering or the opposite: Because prosperity is desired. In either case, this selection on the growth trajectory causes bias. \cite{Redding2015} classify identification strategies for the effect of infrastructure into three broad classes: Planned routes instruments \citep{BaumSnow2007}, historical routes instruments \citep{Duranton2012} or inconsequential units instruments \citep{Berger2017}. The present paper is an innovation in this regard as a \textit{fourth} approach: The emergence of natural infrastructure can be used to study the effect of infrastructure more broadly.

The search for time-variation in effective distances is motivated by the simultaneity of trade and income. If there is trade, then there is growth (in theory) and if there is growth, then this also promotes trade. As a result, there has been discussions on the relationship between trade and growth at least since \cite{ricardo1817principles}. This paper builds on a strand of this debate arising from \cite{Frankel1999}. The insight of \cite{Frankel1999} is that geography (distances specifically) is the source of variation in trade. Using changes in trade predicted by distances can then be used to estimate this otherwise simultaneous relationship. \cite{Rodriguez2001} caution that distance between countries is correlated with other things, which also determine growth, e.g. distance to the equator, institutions, etc. When this is taken into account it is no longer possible to demonstrate an effect. Yet the work of \cite{Frankel1999} still stands as an important inspiration, which motivates the seemingly paradoxical hunt for time variation in distances. The present paper is one such case of time variation in distances.

@@

%Is it relevant to study trade dynamics back then? \cite{FedericoSchulzeVolckart2021} demonstrates how some important important markets where integrated in the 15th century - much before the period studied here.

%\cite{Juhasz2018} demonstrates how French spinning markets faced British competition, (and how this in turn meant that they benefitted from infant-industry-type effects in the Napoleonic Blockade).

% \cite{Redding2008} demonstrates how market access changes account for the development after German reunification, and \cite{Ahlfeldt2015} studies the same for the case of urban reallocation in Berlin.

\section{Historical background}
A fjord is a navigable inlet of water - typically found in Northern Europe and in particular Scandinavia. Fjords act as important natural infrastructure by connecting otherwise remote regions with convenient waterways. The word 'fjord' takes its origin in the Proto-Indo-European 'pertu-', which evolved into many of the words describing roads, access points, or infrastructure more broadly. E.g. 'pertu-' is also the origin 'port', 'per' as in per capita \citep{EtymFjord}. This demonstrates how fjords even in prehistory were perceived as important primarily in terms of their transportation capabilities. The fjords acted as the Scandinavian highways in a time before modern transportation systems. Looking at a map, it is noticeable that many Scandinavian towns and cities, even today, are located on the shore of a fjord.\footnote{Stockholm, Oslo, Gothenburg, Bergen, Stavanger, Odense, Aalborg and Roskilde just to name a few important cities from around Scandinavia, which fit this description.}

The story of the present paper is centred on the so-called Lim-fjord, named after the high concentration of subterranean lime \citep{lexlimfjord2017}. It is located in Northwestern Denmark, more specifically the Northern part of Jutland (the peninsula 'jutting' out of Germany). The region is located conveniently at the heart of Scandinavia, between modern-day Norway and Sweden. Before the 1100s, the Limfjord had both an eastern and a western opening, making it a safer shortcut for ships wanting to avoid the rough sea of Skagerak - the strait between Norway and Denmark. It is no surprise that the Limfjord developed into the main hub for Vikings to gather before sailing west to trade (and pillage) \citep{Rasmussen1966}. In the centre of the Limfjord, the settlement of Aggersborg grew large and would become the location of the largest of a series of so-called ring castles constructed in the 10th century when the Jelling dynasty (of which Harald Bluetooth is probably the most well-known) consolidated power over an emerging Danish state \citep{pedersen2014}. But this would not last. The western opening would close and the local economy would stagnate, as shown by the analysis of the present paper.

>>>>

    \begin{center}
        \LARGE
        \textbf{A Perfect Storm: First-Nature Geography and Economic Development} \\

@@

    \normalsize
    Is geography destiny? What is the role of first-nature geography in determining prosperity? This paper estimates the effect of randomly removing and introducing favorable first-nature geography to a specific region using a difference in difference design. In 1825 a storm introduced a new navigable waterway, bringing trade and prosperity to the otherwise relatively isolated northwestern Denmark. And 700 years prior, the same event happened in reverse, when a previous channel closed up between 1086 and 1208. The elasticity of geography-induced market access is estimated to be 1.6, corresponding to 24 percent population growth within a generation of the event. Demonstrated mechanisms include trade, fertility, fishing, and the rise of manufacturing. The central finding is replicated in reverse in a register of dated archaeological sites. The 1086-1208 closing caused fewer buildings and sites containing coins. The general insight is the same: First-nature geography determines the levels and location of prosperity.

@@

    \footnotesize
    $^*$This project was made possible through generous funding by the Independent Research Fund Denmark (DFF – 6109-00123). I want to thank Paul Sharp, Christian Møller-Dahl, Casper Worm Hansen, Max Schulze, Gregory Clark, James Fenske, Mathias Barding, Torben Johansen, Nadja van't Hoff, Kerstin Enflo, Neil Cummins,- and former PhD student-colleagues (too numerous to mention individually) both at SDU and at LSE for valuable feedback. This project also benefitted from the conscientious research assistance of Andreas Slot Ravnholt. This project was presented at the 2021 and 2022 DGPE workshops, the 2022 EHES conference, the SSEH meeting in 2022, in the LSE Economic History Graduate Seminar series in 2022, the EHS conference in 2023, the End-of-Semester workshop of the Department of Economics 2023 (University of Copenhagen), the Annual CITP conference (Nottingham), the Danish Society of Transport Economics (Copenhagen) and multiple times at the PhD seminar series at the Department of Economics (University of Southern Denmark). I am grateful for the insightful comments and suggestions by participants at these events. This paper also benefitted from improvements suggested by ChatGPT. Important early inspiration for this paper came from the podcast 'Kongerækken' and Poulsen (2019). Finally, I want to thank my fiancee, Maria Fay Courtney Bohr. On our vacation during the COVID-summer of 2020, we stumbled over some fascinating local history resulting in the present paper.  All mistakes are my own. The paper was previously called "A perfect storm and the natural endowments of trade-enabling infrastructure". All code to reproduce all the steps from raw data to the final regressions and the paper itself is available at \url{https://github.com/christianvedels/A_perfect_storm}

@@

\section{Introduction}
In the broadest sense, there are two types of long-run explanations for why some places prosper while others struggle: Institutions \citep{rodrik2004institutions} and geography \citep{Henderson2018satelite}. A predominant consensus suggests that institutions play a primary role, with geography exerting its influence through its impact on the formation of institutions \citep{Acemoglu2001, rodrik2004institutions, Easterly2003, Ketterer2018}. But even among geographical explanations, the evidence to suggest the importance of path-dependence and second-nature geography \citep{Krugman1991, Bleakley2012, Ager2020a} and relegates the role of first-nature geographical fundamentals as being merely important as seeds of future growth and development \citep{Davis2002, Bosker2017, Allen2023}. This paper demonstrates the importance of first-nature geography in its own right. This is done by using a natural experiment that suddenly changed first-nature geography. This in turn caused an increase in trade, population density, and both manufacturing and fishing: A large permanent causal effect of first-nature geography on the level and location of economic activity.

In 1825 a violent storm breached the Agger Isthmus in Northwestern Denmark. An erosion process was initiated and in 1834 a navigable channel had formed. The isthmus had previously separated the Limfjord from the North Sea. But now the region was endowed with convenient access to the open ocean and world markets (see Figure \ref{fig:main_map}). Market towns of the West Limfjord had previously been small and isolated. Now they grew to become relatively important ports of international trade. Similarly, a channel in approximately the same location closed between 1085 and 1208. After the last ice age, Northwestern Denmark was left with a convenient westward channel, used by traders and Viking fleets alike. But it would close due to gradual land-rises - an ongoing process since the last ice age \citep{Christensen2004}. What followed was a long-run stagnation of the West Limfjord region until the geographic circumstances were reversed in 1825.

This paper provides evidence for the effects of these shocks in a difference-in-difference framework. This is done using historical census data, a large database of Northern European historical trade (the Sound Toll Register) and the densities of archaeological findings. It is shown that the 1825 breach led to an increase in population of 24 percent in 1901. To document the mechanisms, I have collected new data on all Danish occupations (13.5 million observations) in the period 1787-1901. The Danish censuses contain a description of each person's occupation. A new machine learning procedure, \textit{OccCANINE}, is applied to these descriptions to convert them into standardized HISCO codes with high accuracy \citep{leeuwen2002hisco, dahl2024breaking}. The HISCO-coded census data is then used to estimate the effect on occupational structure. The key finding is that the channel caused both fishing and manufacturing to grow. An important question is whether the improved geography caused growth or simply reallocation of prosperity. Using age, gender and birthplace information it is shown that the population growth came primarily from fertility - not migration. Denmark of the 19th century was in a post-Malthusian regime, where improved living standards cause fertility \citep{Jensen2022, Galor2011}. Taken together, the evidence of this paper suggests that intrinsic growth is the most likely explanation of the observed effect. That is, improved first-nature geography seemingly causes intrinsic growth - not just a reallocation of growth that would have occurred elsewhere.

For the 11th-century closing, the analysis relies on archaeological data and the closing of the previous channel as an exogenous shock. The time variation comes from the dating of archaeological sites. The geolocation of sites containing coins and buildings is used to construct panels of the probable location of economic activity. A branch within archaeology, Settlement Scaling Theory \citep{Ortman2020}, argues for such approaches to infer historical economic activity, and archaeological evidence has also seen recent application within economics \citep{Davis2002, Barjamovic2019, Bakker2021Phonecians, Allen2023}. In this paper, I develop a novel method to determine the probability that a coin or building finding was left for later archaeological discovery at any given time. Results show that a one percent decrease in market access led to a 13 percent decrease in the probability of coin findings being generated. Denmark of the 11th century is very different from its 19th-century counterpart. But despite differences in technology, culture, institutions, religion, and most other aspects thought important for economic development, geomorphology played a similar role in the 12th century as it did in the 19th century, providing evidence of (temporal) external validity of the effect of first-nature geography on prosperity.

A separate but related contribution of this paper is in understanding the benefits of infrastructure. Some of the most expensive and extensive projects that a society can undertake are infrastructure projects. Infrastructure changes distances between people. Infrastructure alters first-nature geography. However, infrastructure investments are often endogenous to outcomes of interest.  If a dramatic increase in e.g. population density is observed following infrastructure, it is difficult to identify whether this is because it was just constructed in an already thriving place. There are different standardized identification strategies commonly applied in the literature \citep{Redding2015}. This paper can be seen as an entirely new way of studying infrastructure change by relying on changes in the natural endowments of it.

@@

This paper proceeds as follows. Section 2 outlines this paper's relation to the literature. Section 3 presents the historical background. Section 4 presents the overall empirical strategy. Section 5 presents a general overview of the data used. The remaining sections demonstrate empirical results. Section 6 demonstrates how the shock caused an increase in trade. Section 7 shows how the channel caused the population to grow. Section 8 covers mechanisms. Section 9 demonstrates the effect of the closed-up channel in the 1100s. Section 10 concludes.

@@

This paper enters a debate about the fundamental determinants of prosperity. Is geography or are institutions the fundamental cause? \cite{Acemoglu2001} argue that extractive institutions set up by colonizing Europeans are partially to blame for the current low-income status of many African nations. \cite{rodrik2004institutions} argue that geography plays a limited role if institutions are kept constant. \cite{Henderson2018satelite} use satellite image data to show that geography through trade and agriculture can explain 47 percent of the worldwide variation in economic activity. \cite{Ketterer2018} find that first nature does play a role but institutions are far more important even when a comparison is made within European regions. It would be natural to get the impression, that geography is simply unimportant in its own right. What I demonstrate in this paper is the contrary.

Before introducing this in more detail, it is useful to provide some clarity regarding some terms already used and provide a simple taxonomy of geography and economics. \cite{cronon1992nature} introduced the distinction between first- and second-nature geography, where first-nature geography is everything determined independently of humans, and second-nature geography is all of the geography caused by humans and their interactions. This has since become the standard framework for understanding the interplay of economics and geography in broad terms \citep{Caruana-Galizia_Okubo_Wolf_2021}. Among first-nature geography, we can distinguish between:
\begin{enumerate}
    \item \textit{Things in/on the soil/land}: Soiltype, natural resources, rainfall, etc.
    \item \textit{Geomorphology}:\footnote{Also often called 'physical geography'. But this term is confusing. A bridge is very physical. But it is not what is meant by 'physical geography'. A land-bridge on the other hand is 'physical geography'. Using the term Geomorphology circumvents this confusion.} Waterways, lakes, mountains, etc.
\end{enumerate}

A huge literature is dedicated to second-nature geography in terms of infrastructure (covered below). Among first-nature explanations of the 'what's in the soil'-type, \cite{Nunn2011a, HeavyPlough2016, Dalgaard2020} demonstrate that variation in the productivity for food (the potato, the heavy plow, and fish stock respectively) of a particular place caused increased population density. But 'what's in/on the soil'-first-nature also affects other outcomes.  \cite{Decet2023WaterWars} show how a lack of rainfall is associated with an increased level of conflict in modern-day Africa. \cite{WinnersAndLosers2022} demonstrates how land productivity caused inequality, and \cite{clark2014} demonstrate the influence on literacy. Of course, the ground might be endowed with other determinants of growth like coal \citep{ORourkeCoal2021} or peat \citep{CoalPeatPaper}.

On the other hand, first-nature \textit{geomorphology} explanations are more sparsely covered with \cite{Diamond1997, Allen2023, Matranga2024} - all related to the neolithic revolution and the emergence of the first civilizations - as the most notable exceptions. Furthermore \cite{Henderson2018satelite} and \cite{Bosker2017} demonstrate that \textit{geomorphology} is correlated with prosperity. Of course, this is often based on the comparison between places with and without certain geomorphological features, which makes the causality contingent on the degree to which you can trust their method for controlling other differences between these places. Achieving a clean identification remains a formidable challenge. Consequently, a survey by \cite{Bosker2022} calls for the exploration of natural experiments to gain insights into the causal effects of first-nature \textit{geomorphology} on population growth. The present paper tries to fill that gap.

All of this is related to a central result by \cite{Davis2002}: Even very large disruptions of city populations (the bombs over Hiroshima and Nagasaki at the end of WWII) do not change the long-run outcomes. \cite{Brakman2004Bombs} replicate a similar result for the bombing campaigns in Germany. This result has two types of potentially overlapping explanations: It is either because \textit{first-nature geography} is the \textit{fundamental} determinant of city population for a given state of the economy, or it is because of path dependence.\footnote{\cite{CraftsWolf2014} shows a case for this: Cotton mills grew where the first-nature geography, in the form of available water power, was prevalent. When this was replaced with steam power, the location was persistent due to an interaction of first and second-nature geography - some first-nature geography (water access) is useful for both water power and steam-powered mills and the investments in buildings and surrounding infrastructure (second-nature), might as well be made to good use rather than reallocating it.} In both cases, city sizes will tend to gravitate towards their long-run trend despite disruptions. But it is hard to disentangle the causes. However, if the location of prosperity is fundamentally determined by first-nature geography, then a shock to this should have a permanent rather than temporary effect. This is what I demonstrate: First-nature geography does indeed have a permanent impact on the location of prosperity.

That is not to say, that path dependence does not play a role. \cite{Bleakley2012} convincingly document a case of path dependence. Places that used to be in favorable geography continued to be urban centers even when the geographical features that made them so became irrelevant. An implication of \cite{Krugman1991} is that small initial first-nature geographical differences can accumulate into very different outcomes via second-nature geographical interactions \citep{Caruana-Galizia_Okubo_Wolf_2021}, which is consistent with the observation of \cite{Bleakley2012}. The present paper documents fairly large and permanent effects within a generation when first nature changes, which firmly demonstrates an important role of the \textit{first-nature fundamentals} explanation of the location of economic activity. When favorable geomorphology in the Limfjord region deteriorated, so did the local economy, and when it reverted 700 years later, the economy (and population) increased dramatically. It is not \textit{just} path dependence.\footnote{Two more related publications are \cite{Ager2020a} and \cite{Cermeno2019}. \cite{Ager2020a} show how a natural disaster can change relative city sizes. And on the other hand \cite{Cermeno2019} show that institutions can create towns in otherwise unfavorable geography.}

The central natural experiment used in this paper is related to \citep{Seror2020Random}. He uses the randomly altered courses of the Yellow River in China to measure the effects of waterways on population growth. The paper finds that market access led to increased population, increased taxes, and in turn reallocation of elites - presumably to avoid the taxes. \citep{Allen2023} uses river shifts and archaeological data in what is today Iraq to demonstrate that river shifts are associated with state formation. In comparison to these two papers, the relatively rich Danish data and the unique natural experiment, allow us to exploit both the 11th- and 19th-century effects of the same event, and especially the 19th-century data, combined with detailed historical accounts, makes it possible to explore mechanisms.

Geomorphology affects market access. That is the fundamental mechanism here. As such, we can also compare the present paper to the literature on market access and infrastructure. Infrastructure is second-nature geography, but a special case: It is a deliberate change of the first-nature \textit{geomorphology}. Railways, roads, bridges, canals - even the internet - all change the effective distance between people, which used to be different because of the geomorphology. In general, a large number of papers show how infrastructure shapes everything from land value, health, industrialization, manufacturing and the persistent location of modern-day economic activity \citep{Donaldson2016, Berger2017, Berger2019a, Hornbeck2019, Bogart2019, Zimran2020a, Bogart2022, Hornbeck2024, Hornung2015}. Of course, a central worry is the endogeneity of infrastructure. Infrastructure is often built because a place is already prospering or the opposite: Because prosperity is desired. In either case, this selection on the expected growth trajectory causes bias. \cite{Redding2015} classify identification strategies for the effect of infrastructure into three broad classes: Planned routes instruments \citep{BaumSnow2007}, historical routes instruments \citep{Duranton2012} or inconsequential units instruments \citep{Berger2017}. The present paper is an innovation in this regard as a \textit{fourth} approach: The emergence of natural infrastructure.

The search for time-variation in effective distances is also motivated by the simultaneity of trade and income. If there is trade, then there is growth (in theory) and if there is growth, then this also promotes trade. As a result, there has been discussions on the relationship between trade and growth at least since \cite{ricardo1817principles}. This paper builds on a strand of this debate arising from \cite{Frankel1999}. The insight of \cite{Frankel1999} is that geography (distances specifically) is the source of variation in trade. Using changes in trade predicted by distances can then be used to estimate this otherwise simultaneous relationship. \cite{Rodriguez2001} caution that distance between countries is correlated with other things, which also determine growth, e.g. distance to the equator, institutions, etc. When this is taken into account it is no longer possible to demonstrate an effect. Yet the work of \cite{Frankel1999} still stands as an important inspiration, which motivates the seemingly paradoxical hunt for time variation in distances. The present paper is one such case of time variation in distances.

@@

\section{Historical background}
A fjord is a navigable inlet of water - typically found in Northern Europe and in particular Scandinavia. Fjords act as important natural infrastructure by connecting otherwise remote regions with convenient waterways. The word 'fjord' takes its origin in the Proto-Indo-European 'pertu-', which evolved into many of the words describing roads, access points, or infrastructure more broadly. E.g. 'pertu-' is also the origin 'port', 'per' as in per capita \citep{EtymFjord}. This demonstrates how fjords even in prehistory were perceived as important, primarily in terms of their transportation capabilities. The fjords acted as the Scandinavian highways in a time before modern transportation systems. Looking at a map, it is noticeable that many Scandinavian towns and cities, even today, are located on the shore of a fjord.\footnote{Stockholm, Oslo, Gothenburg, Bergen, Stavanger, Odense, Aalborg and Roskilde just to name a few important cities from around Scandinavia, which fit this description.}

The story of the present paper is centered on the so-called Lim-fjord, named after the high concentration of subterranean lime \citep{lexlimfjord2017}. It is located in Northwestern Denmark, more specifically the Northern part of Jutland (the peninsula 'jutting' out of Germany). The region is located conveniently at the heart of Scandinavia, between modern-day Norway and Sweden. Before the 1100s, the Limfjord had both an eastern and a western opening, making it a safer shortcut for ships wanting to avoid the rough sea of Skagerak - the strait between Norway and Denmark. It is no surprise that the Limfjord developed into the main hub for Vikings to gather before sailing west to trade (and pillage) \citep{Rasmussen1966}. In the center of the Limfjord, the settlement of Aggersborg grew large and would become the location of the largest of a series of so-called ring castles constructed in the 10th century when the Jelling dynasty (of which Harald Bluetooth is probably the most well-known) consolidated power over an emerging Danish state \citep{pedersen2014}. But this would not last: The western opening would close and the local economy would stagnate, as shown by the analysis of the present paper.

The last historical evidence of the use of a western opening was in 1085. King Canute IV gathered his fleet in the Limfjord, to sail west and uphold his ‘claim’ to the English throne after his great uncle Canute II ('the great') who had been King of the so-called North Sea empire consisting of England, Denmark, Norway and parts of Sweden \citep{Spejlborg2012}. The sailors and soldiers rebelled against Canute IV, the fleet never left Denmark and he was killed when seeking refuge in Saint Alban’s church in the town of Odense in 1085.\footnote{The very place where I wrote this paper.} He was later sainted for this and earned the nickname ‘the holy’ \citep{Pajung2012}. This event marks the last attempt at gathering a Viking fleet in Denmark to sail west. Never again would Vikings be able to use the western channel. Never again would Danish Vikings try to conquer England \citep{Roesdahl2009}.

The next time this channel shows up in history is in the writings of the early historian Saxo \cite{saxo}. Saxo reports in the past tense that there 'used to' be a convenient Western channel.\footnote{\cite{saxo} book XIII, section 5} There is some uncertainty surrounding the timing of the writings of Saxo. But it is known that it was around the year 1200, and the historian, \cite{Mortensen2018}, notes that the year 1208 is most likely. Therefore, it can be determined that the western channel probably closed sometime between 1085 and 1208. This approximate timing is supported by geological surveys. By drilling through the layers of soil and deposits, geologists have demonstrated a change in the levels of salinity consistent with a closing channel around this time \citep{Christensen2004}.
